Columnar VCF using Zarr

### 2024-03-28 paper.tex
Stuff about vcf2zarr

Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is widely
used and is being incorporated, e.g., as a standard compression
mechanism for HTTP traffic and to decompress Linux kernel images
during boot.

>>>>

Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is widely
used and is being incorporated, e.g., as a standard compression
mechanism for HTTP traffic and to decompress Linux kernel images
during boot.
% LZ4 is also notable - any citations for this?

\subsection{The vcfzarr specification}
The vcfzarr specification is a direct mapping from the VCF data model 
to a columnar binary format, expressed using Zarr. [A HIGH LEVEL 
DESCRIPTION OF THE SPEC, FOCUSING ON RATIONALE BEHIND DECISIONS]

The vcfzarr specification can represent anything described by BCF
(which is somewhat more restrictive than BCF) except for two corner
cases related to the encoding of missing data. Firstly, vcfzarr does
not distinguish between a field that is not present and one that 
is present but contains missing data. [FOR EXAMPLE...]
Because of the use of -1 to represent missing data and -2 to 
represent dimension padding for integers, these values 
cannot be stored. In practise this doesn't seem to be much of 
an issue (we have not found a real VCF that contains negative 
integers). If -1 and -2 need to be stored, a float field
can be used without issues.

\subsection{vcf2zarr}
Converting VCF to Zarr format at Biobank scale is challenging. This main issue is 
around memory usage: although we can view each record in the VCF one-by-one,
we must buffer a full chunk (10,000 variants is reasonable choice) 
in the variants dimension for each of the arrays. 
For VCFs with many FORMAT fields and large numbers of samples this can
easily require tens of gigabytes of RAM per worker, which makes 
parallelism troublesome. Reading the VCF multiple times for different fields
is possible, but would be prohibitively slow for multi-terabyte VCFs.

The \texttt{vcf2zarr} utility solves this problem by first converting 
the VCF data (which can be split across many files) into an Intermediate
Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
of VCFs and stores the data in per-field chunks (by default, 64MiB).
In this way, the memory required is bounded by the size of single record 
BCF record in htslib (plus the size of a decoded bgzf record
[FIXME what is the correct terminology here?]), plus 64MiB for each 
field.  The Intermediate Columnar Format stores the data for each 
field spread across multiple partitions of the data (so that 
multiple parts of the VCF can be read in parallel), and then 
fixed size compressed chunks within those partitions. It also stores information
about the number of records in each partition and chunk of a given
field, so that the record at a particular index can be retrieved efficiently 
later. Summaries such as the minimum and maximum value 
of each field are also maintained, to aid choice of data types later.
A set of VCF files can be converted to intermediate columnar 
format in parallel on a single machine using a given number 
of worker processes using the \texttt{explode} command,
or can be distributed across a cluster using the 
\texttt{dexplode-init},
\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.

Once the VCF data has converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. Alhough there is a one-to-one mapping between the VCF
fields and the Zarr arrays, there is considerable flexibility 
in how the data is stored, for example in integer width
and compression and filter settings.
By default vcf2zarr chooses integer type sizes based 
on the range of values seen in the intermediate column format
and reasonable compressor defaults, but these can be
altered by generating a schema in JSON format and editing it.
Encoding a given column (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from the intermediate columnar format. Once the buffer 
is full, the chunks in the samples dimension are compressed
and written to file. [Something about parallelism and distributing]
